A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
fifteen pages!

\begin{rep}\paragraph{Related Work.} Despite\end{rep}%
\begin{conf}Despite\end{conf} its age,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in both types of provers \cite{kersani-peltier-2013,wand-weidenbach-201x,reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus \cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

@@

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or a node of the form
$\MU x.\: \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.

%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.

%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$\begin{rep},
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can represent the same value while not being $\alpha$-equivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\enskip %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\enskip %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again to refine the equivalence classes,
yielding
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{conf}%
$\Val \ec{y} \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})$.
\end{conf}%
\begin{rep}
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
\end{rep}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule can finally derive $x \teq y$.
\xend
\end{examplex}

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
(Section~\ref{sec:the-theory-of-co-datatypes}) and must therefore be equal. Both
the finiteness assumption on the datatype in \rn{Split} and the singleton
constraint on the codatatype in \rn{Single} can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types. This step relies on the assumption that all types
in $\Nondata$ are infinite.

>>>>

\begin{rep}\paragraph{Related Work.} Despite\end{rep}%
\begin{conf}Despite\end{conf} its age,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related is the automatic structural
induction in both types of provers \cite{kersani-peltier-2013,wand-weidenbach-201x,reynolds-kuncak-2015}.
%
Datatypes are explicitly supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also find models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus
\begin{conf}\cite{kozen-1983}\end{conf}%
\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}\end{rep}%
.

@@

\kern-1.5pt %% TYPESETTING hack

@@

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or a node of the form
$\MU x.\: \const{C}( \vec t\, )$ for some constructor $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. \hfill It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. \hfill The
\goodbreak %%% TYPESETTING: horrible hacks (including two hfills)
\noindent
notation $t \aequiv u$, for $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\enskip %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\enskip %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{conf}%
$\Val \ec{y} \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})$.
\end{conf}%
\begin{rep}
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
\end{rep}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
\rn{Unique} can finally derive $x \teq y$.
\xend
\end{examplex}

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
%(Section~\ref{sec:the-theory-of-co-datatypes})
and must therefore be equal. Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types. This step relies on the assumption that all types
in $\Nondata$ are infinite.
